IFA Comparison of Noncognitive Skills \\ BCS and MCS

### 2018-02-23 codes/R/cohorts_nc.Rnw
selection of score model in separate chunk

For BCS, I augment the 20-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.
\item Is squirmy or fidgety.
\item Often destroys own or others’ belongings.
\item Frequently fights other children.
\item Not much liked by other children.
\item Often worried, worries about many things
\item Tends to do things on his/her own – rather solitary.
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\item Frequently bites nails or fingers.
\item Is often disobedient.
\item Cannot settle to anything for more than a few moments.
\item Tends to be fearful or afraid of new things or new situations.
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.
\item[A] Complains of headaches.
\item[B] Complains of stomach-ache or has vomited.
\item[C] Complais of biliousness.
\item[D] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).
\end{enumerate}
The Rutter items are rated on three levels: Does not apply, Somewhat applies, Certainly applies. Since they are all indicators of low skills, I encode all of them in reverse, i.e. ``Certainly applies" = 0, ``Somewhat applies" = 1, ``Does not apply" = 2. The last 4 items are rated on 4 levels: Never in the last 12 months, less than once a month, at least once a month, at least once a week. I recode these into binary indicators, with \emph{Never} and \emph{Less than once a month}=1 and zero otherwise.

@@

The five-factor structure at the basis of the full SDQ has been validated in many contexts \citep{Stone2010}; however, lower-dimensional structures have been also suggested \citep{Dickey2004}, and recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter \citep{Goodman2010}. To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 12-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample. Different methodologies for quantifying the optimal number of factors produce inconclusive results, but both Velicer MAP and very simple structure (VSS) seem to indicate that a two-factor structure is suitable. An EFA model with two factors, employing the usual cutoff of .3 for the loadings, delivers a neat and sensible separation between items, as shown in the second column of \autoref{tab:comm}. Thus, our two-factor structure seems justifiable on both theoretical and empirical grounds.

@@

\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{cj} + u_{cji} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{cj} + u_{cji} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

@@

For each item $i$, we observe either a binary or ordinal (3-categories) manifestation $X_{cji}$. Given \eqref{eq:propens}, the observed outcome as a function of the propensities $X^*$ can be defined as a function of additional threshold parameters $\tau$, as follows:
$$X_{cji} = s \qquad \text{if} \; \tau_{s,ci} \leq X^*_{cji} < \tau_{s+1,ci} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ci}=-\infty$ and $\tau_{3,ci}=+\infty$.
The probability of observing each category for a binary item takes values:
\begin{equation}
X_{cji}=
\begin{cases}
\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
\qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
\end{cases}
\end{equation}

@@

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}\bm{\Psi}=\bm{I}.$$

@@

<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
@

<<FA_FIT, dependson='FA_INVARIANCE'>>=
@

<<FA_FINAL, dependson='FA_INVARIANCE'>>=
@

<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
@

@@

<<FA_SCORES, dependson='FA_FINAL'>>=
@

>>>>

For BCS, I augment the 19-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.
\item Is squirmy or fidgety.
\item Often destroys own or others’ belongings.
\item Frequently fights other children.
\item Not much liked by other children.
\item Often worried, worries about many things
\item Tends to do things on his/her own – rather solitary.
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\item Frequently bites nails or fingers.
\item Is often disobedient.
\item Cannot settle to anything for more than a few moments.
\item Tends to be fearful or afraid of new things or new situations.
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.
\item[A] Complains of headaches.
\item[B] Complains of stomach-ache or has vomited.
\item[C] Complais of biliousness.
\item[D] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).
\end{enumerate}
The Rutter items are rated on three levels: Does not apply, Somewhat applies, Certainly applies. Since they are all indicators of low skills, I encode all of them in reverse, i.e. ``Certainly applies" = 0, ``Somewhat applies" = 1, ``Does not apply" = 2. The last 4 items are rated on 4 levels: Never in the last 12 months, less than once a month, at least once a month, at least once a week. I recode these into binary indicators, with \emph{Never} and \emph{Less than once a month}=1 and zero otherwise.

@@

The five-factor structure at the basis of the full SDQ has been validated in many contexts \citep{Stone2010}; however, lower-dimensional structures have been also suggested \citep{Dickey2004}, and recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter \citep{Goodman2010}. To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 11-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample. Different methodologies for quantifying the optimal number of factors produce inconclusive results, but both Velicer MAP and very simple structure (VSS) seem to indicate that a two-factor structure is suitable. An EFA model with two factors, employing the usual cutoff of .3 for the loadings, delivers a neat and sensible separation between items, as shown in the second column of \autoref{tab:comm}. Thus, our two-factor structure seems justifiable on both theoretical and empirical grounds.

@@

\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

@@

For each item $i$, we observe either a binary or ordinal (3-categories) manifestation $X_{cji}$. Given \eqref{eq:propens}, the observed outcome as a function of the propensities $X^*$ can be defined as a function of additional threshold parameters $\tau$, as follows:
$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
The probability of observing each category for a binary item takes values:
\begin{equation}
X_{cji}=
\begin{cases}
\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
\qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
\end{cases}
\end{equation}

@@

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$

@@

<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
# estimate models of different levels of invariance
@

<<FA_FIT, dependson='FA_INVARIANCE'>>=
# assemble fit indices into matrices
@

<<FA_FINAL, dependson='FA_INVARIANCE'>>=
# select final model set
@

<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
# get parameters of final model
@

@@

<<FA_SCORES, dependson='FA_FINAL'>>=
# produce scores for different models
@

<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
# assemble and compare scores across different models
@

<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
# select model to use for inequality analysis
@
